# Geology map (experiment)

```{r}
#| label: setup
#| include: false

source(here::here("scripts/setup-libraries.R"))
source(here("scripts/setup-data-prep.R"))

```

```{r}
sf_use_s2(FALSE)

xx <-  st_read(here("data/map/NorthCarolinaGeologicMap/NorthCarolinaGeologicMap/"))

```

## Test geologic map shapefile

<https://files.nc.gov/ncdeq/Energy%20Mineral%20and%20Land%20Resources/Geological%20Survey/Assorted%20maps/NorthCarolinaGeologicMap.zip> and <https://www.deq.nc.gov/energy-mineral-and-land-resources/geological-survey/assorted-maps/info-1985-state-geologic-map/download> downloaded from <https://www.deq.nc.gov/about/divisions/energy-mineral-and-land-resources/north-carolina-geological-survey/ncgs-publications/statewide-geologic-maps-north-carolina> on 2024-02-24


x2 does the geographic join OK to the shape of nc_counties but then is not picking up the features from xx

```{r fig.height=8, fig.width=12}
#| label: fig-geology-four-counties-x2
#| fig-cap: "x2"
#| fig-height: 8
#| fig-width: 12
#| column: screen-right
#| eval: false
##| warning: false
##| message: false

x2 <- st_join(
  nc_counties,
  #st_as_sf(nc_counties_border_union),
  st_transform(xx, prj),
  join = st_intersects)

x2 |>
ggplot() +
  geom_sf(#data = x2,
          aes(fill = TYPE)
  )
```

<br>

x3 doesn't seem to be picking up the intersects. But it can display xx's features OK,
I cannot tell what "aoi" (area of interest) does in st_transform(). Doesn't seem to be a bounding box.

```{r fig.height=8, fig.width=12}
#| label: fig-geology-four-counties-x3
#| fig-cap: "x3"
#| fig-height: 8
#| fig-width: 12
#| column: screen-right
#| eval: false
##| warning: false
##| message: false

x3 <- st_join(#nc_counties,
  st_transform(xx, prj,
               aoi = st_bbox(nc_counties)), ## |> st_set_bbox(st_bbox(nc_counties))
  st_as_sf(nc_counties_border_union),
  join = st_overlaps) 

ggplot() +
  geom_sf(data = x3,
          aes(fill = BELT)
  )

ggplot() +
  geom_sf(data = x3,
          aes(fill = TYPE)
  )

ggplot() +
  geom_sf(data = x3,
          aes(fill = TYPE)
  )

ggplot() +
  geom_sf(data = x3,
          aes(fill = FORMATION),
          show.legend = FALSE
  )
```

<br>

I can't tell what x4 is doing. Not the right thing!

```{r fig.height=8, fig.width=12}
#| label: fig-geology-four-counties-x4
#| fig-cap: "x4"
#| fig-height: 8
#| fig-width: 12
#| column: screen-right
#| eval: false

x4 <- st_intersection(st_transform(xx, prj,
                                   st_as_sf(nc_counties_border_union)
                                   )
)

ggplot() +
  geom_sf(data = x4,
          aes(fill = TYPE)
  )

```

<br>


```{r fig.height=8, fig.width=12}
#| label: fig-geology-four-counties-x4a
#| fig-cap: "x4"
#| fig-height: 8
#| fig-width: 12
#| column: screen-right
#| eval: false

x4 <- st_intersection(st_transform(xx, prj,
                                   st_as_sf(nc_counties_border_union)
                                   )
)

x3 <- st_join(#nc_counties,
  st_transform(xx, prj,
               aoi = st_bbox(nc_counties)), ## |> st_set_bbox(st_bbox(nc_counties))
  st_as_sf(nc_counties_border_union),
  join = st_overlaps) 

x5 <- st_intersection(st_transform(xx, prj),
                      nc_counties_border_union)

# datums and ESPG codes are different, so definitely need st_transform()
st_crs(xx)
st_crs(nc_counties_border_union)

# bbox is correct and the same (intersect seems to be right)
st_bbox(x5)
st_bbox(nc_counties_border_union)
```


```{r fig.height=8, fig.width=12}
#| label: fig-geology-four-counties-x5
#| fig-cap: "x5"
#| fig-height: 8
#| fig-width: 12
#| column: screen-right

x5 <- st_intersection(st_transform(xx, prj),
                      nc_counties_border_union)

ggplot() +
  geom_sf(data = x5,
          aes(fill = TYPE),
          color = NA
  ) +
  geom_sf(data = nc_counties,
          fill = NA,
          linewidth = 1) +
  geom_sf(data = nc_rivers,
          color = "blue") +
  geom_sf(data = nc_waterbodies,
          color = "blue", fill = "blue")

ggplot() +
  geom_sf(data = x5,
          aes(fill = GROUP_),
          color = NA
  ) +
  geom_sf(data = nc_counties,
          fill = NA,
          linewidth = 1) +
  geom_sf(data = nc_rivers,
          color = "blue") +
  geom_sf(data = nc_waterbodies,
          color = "blue", fill = "blue")

ggplot() +
  geom_sf(data = x5,
          aes(fill = BELT),
          color = NA
  ) +
  geom_sf(data = nc_counties,
          fill = NA,
          linewidth = 1) +
  geom_sf(data = nc_rivers,
          color = "blue") +
  geom_sf(data = nc_waterbodies,
          color = "blue", fill = "blue")

```

<br>

```{r fig.height=8, fig.width=12}
#| label: fig-geology-four-counties-x6
#| fig-cap: "x6"
#| fig-height: 8
#| fig-width: 12
#| column: screen-right

geology_data <- st_intersection(st_transform(xx, prj),
                      nc_counties_border_union)

ggplot() +
  geom_sf(data = x5,
          aes(fill = GROUP_),
          color = NA
  ) +
  geom_sf(data = nc_rivers, color = "dodgerblue", fill = "dodgerblue",
          linewidth = 0.25) +
  geom_sf(data = nc_waterbodies, color = "dodgerblue", fill = "dodgerblue",
          linewidth = 0.25) +
  geom_sf(data = nc_counties, color="black",
          fill = NA, size = 0.25) +
  geom_sf(data = nc_counties_border_union, color="black",
          fill = NA, size = 1) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.9, 0.8)
        ) +
  labs(
    title = "Add title",
    # subtitle = glue("{comma(n_marks_counties)} marks in {n_counties} NC counties"),
    caption = my_caption
  )


```


<br>

```{r fig.height=8, fig.width=12}
#| label: fig-geology-four-counties-x7
#| fig-cap: "x7"
#| fig-height: 8
#| fig-width: 12
#| column: screen-right

geology_data <- st_intersection(st_transform(xx, prj),
                      nc_counties_border_union)

ggplot() +
  geom_sf(data = x5,
          aes(fill = TYPE),
          color = NA
  ) +
  geom_sf(data = nc_rivers, color = "dodgerblue", fill = "dodgerblue",
          linewidth = 0.25) +
  geom_sf(data = nc_waterbodies, color = "dodgerblue", fill = "dodgerblue",
          linewidth = 0.25) +
  geom_sf(data = nc_counties, color="black",
          fill = NA, size = 0.25) +
  geom_sf(data = nc_counties_border_union, color="black",
          fill = NA, size = 1) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.9, 0.8)
        ) +
  labs(
    title = "Add title",
    # subtitle = glue("{comma(n_marks_counties)} marks in {n_counties} NC counties"),
    caption = my_caption
  )


```

<br>

```{r fig.height=8, fig.width=12}
#| label: fig-geology-four-counties-x8
#| fig-cap: "x8"
#| fig-height: 8
#| fig-width: 12
#| column: screen-right

xx_names <- distinct(xx, Name) |>
  arrange(Name) |>
  rownames_to_column(var = "key")

geology_data <- st_intersection(st_transform(xx, prj),
                      nc_counties_border_union) |>
  inner_join(xx_names, by = "Name")

ggplot() +
  geom_sf(data = geology_data,
          aes(fill = key),
          color = NA
  ) +
  geom_sf(data = nc_rivers, color = "dodgerblue", fill = "dodgerblue",
          linewidth = 0.25) +
  geom_sf(data = nc_waterbodies, color = "dodgerblue", fill = "dodgerblue",
          linewidth = 0.25) +
  geom_sf(data = nc_counties, color="black",
          fill = NA, size = 0.25) +
  geom_sf(data = nc_counties_border_union, color="black",
          fill = NA, size = 1) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.95, 0.8)
        #legend.position = "bottom"
        ) +
  guides(fill = guide_legend(ncol = 2)) +
  labs(
    title = "Rock types",
    # subtitle = glue("{comma(n_marks_counties)} marks in {n_counties} NC counties"),
    # fill = "     ",
    fill = "Key",
    caption = my_caption
  )

```

<br>

```{r}

geology_data |>
  st_drop_geometry() |>
  mutate(key = as.numeric(key)) |>
  arrange(key) |>
  replace_na(list(Name = "Other")) |>
  distinct(key, Name) |>
  rename(`Rock Type` = Name) |>
  gt()

```

<br>

```{r fig.height=8, fig.width=12}
#| label: fig-geology-four-counties-diabase
#| fig-cap: "x8"
#| fig-height: 8
#| fig-width: 12
#| column: screen-right

xx |>
  filter(FORMATION == "Diabase") |>
  ggplot() +
  geom_sf(fill = "firebrick",
          alpha = 0.4) +
  # geom_sf(data = geology_data,
  #         aes(fill = key),
  #         color = NA
  # ) +
  geom_sf(data = nc_rivers, color = "dodgerblue", fill = "dodgerblue",
          linewidth = 0.25) +
  geom_sf(data = nc_waterbodies, color = "dodgerblue", fill = "dodgerblue",
          linewidth = 0.25) +
  # geom_sf(data = nc_counties_all_tigris, color="black",
  #         fill = NA, size = 0.25) +
  geom_sf(data = nc_counties_border_union, color="black",
          fill = NA, size = 1) +
  geom_sf(data = nc_counties, color="black",
          fill = NA, size = 1) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.95, 0.8)
        #legend.position = "bottom"
        ) +
  guides(fill = guide_legend(ncol = 2)) +
  labs(
    title = "Diabase",
    # subtitle = glue("{comma(n_marks_counties)} marks in {n_counties} NC counties"),
    # fill = "     ",
    #fill = "Key",
    caption = my_caption
  )

```