# FY3210 in context

```{r}
#| label: setup
#| include: false

source(here::here("scripts/setup-libraries.R"))
source(here("scripts/setup-data-prep.R"))

```

FY3210 is one of over 2900 marks in these four counties.

## Survey mark network

```{r fig.height=8, fig.width=12}
#| label: fig-mark-locations
#| fig-cap: "Survey mark 'network'"
#| fig-height: 8
#| fig-width: 12
#| column: screen-right
#| warning: false

ggplot() +
  geom_sf(data = nc_counties, color="black",
          fill = NA, size = 0.25) +
  geom_sf(data = nc_marks_counties |>
            filter(county %in% c("Orange", "Chatham", "Durham", "Wake")),
          #aes(color = ortho_ht), #* 3.28084 convert to ft
          size = 1, color = "white",
          show.legend = TRUE
          ) +
  geom_sf(data = nc_marks_counties |>
            filter(county %in% c("Orange", "Chatham", "Durham", "Wake")),
          aes(color = ortho_ht), #* 3.28084 convert to ft
          size = 0.5,
          show.legend = TRUE
          ) +
  geom_sf(data = nc_counties_border_union, color="black",
          fill = NA, size = 1) +
  scale_color_gradientn(colors = pal_dem_screen$hex,
                        # values = scales::rescale(pal_dem_screen$limit),
                        # limit = range(pal_dem_screen$limit),
                        # values = my_values,
                        limit = my_range,
                        na.value = "transparent",
  ) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.9, 0.8)) +
  labs(
    title = "Survey mark 'network'",
    subtitle = glue("{comma(n_marks_counties)} marks in {n_counties} NC counties"),
    caption = my_caption
  )

```

<br>

The geographical distribution of marks looks like a road network, because at least in these counties, survey marks *are* mostly located on or near roads.^[@fig-counties-osm includes Open Street Map data for roads of type "motorway", "trunk", "primary", "secondary", "tertiary", and "road".<https://wiki.openstreetmap.org/wiki/Key:highway> ]

```{r fig.height=9, fig.width=13}
#| label: fig-counties-osm
#| fig-cap: "Survey marks mostly follow roads"
#| fig-height: 9
#| fig-width: 13
#| column: screen-right
#| message: false

ggplot() +
  geom_spatraster(data = elev_nc_counties,
                  alpha = 0.7) +
  geom_sf(data = nc_rivers, color = "dodgerblue", fill = "dodgerblue",
          linewidth = 0.25) +
  geom_sf(data = nc_waterbodies, color = "dodgerblue", fill = "dodgerblue",
          linewidth = 0.25) +
  geom_sf(data = nc_counties, color="black",
          fill = NA, size = 0.25) +
  geom_sf(data = nc_marks_counties,
          #aes(color = ortho_ht), #* 3.28084 convert to ft
          size = 0.75, color = "black", 
          # size = 1, color = "white",
          show.legend = TRUE
  ) +
  geom_sf(data = nc_marks_counties,
          aes(color = ortho_ht), #* 3.28084 convert to ft
          size = 0.5,
          show.legend = TRUE
          ) +
  geom_sf(data = nc_counties_border_union, color="black",
          fill = NA, size = 1) +
  geom_sf(data = streets$osm_lines,
          inherit.aes = FALSE,
          color = "black",
          #size = 0.05,
          linewidth = 0.075,
          alpha = 0.7) +
  scale_fill_gradientn(colors = pal_dem_screen$hex,
                       values = scales::rescale(pal_dem_screen$limit),
                       limit = range(pal_dem_screen$limit),
                       na.value = "transparent",
  ) +
  scale_color_gradientn(colors = pal_dem_screen$hex,
                        values = scales::rescale(pal_dem_screen$limit),
                        limit = range(pal_dem_screen$limit),
                        na.value = "transparent",
  ) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.9, 0.8)
        ) +
  labs(
    title = "Survey marks mostly follow roads",
    subtitle = glue("{comma(n_marks_counties)} marks in {n_counties} NC counties"),
    caption = my_caption,
    color = "Elevation (m)",
    fill = "Elevation (m)"
  )

```

<br>

## Many kinds of marks

There are many kinds of marks (@sec-mark-types). The most common are marks determining horizontal location: Transverse Station marks and there predecessors Horizontal Control Disks, Reference Marks, and Boundary Marks. Vertical location marks include Bench Marks and Vertical Control Marks. There are many times more marks focused on horizontal location than elevation. Less common mark types are shown in @fig-mark-types-last-recovery.

```{r, fig.height=8, fig.width=12}
#| label: fig-mark-types-first-and-last-recovery
#| fig-cap: "Survey and bench mark types: first and last recovery"
#| fig-height: 8
#| fig-width: 12
#| column: page-right

data_for_plot <- nc_marks_counties |>
  st_drop_geometry() |>
  filter(!is.na(last_recv),
         !is.na(first_recv)) |>
  mutate(first_recv = as.numeric(str_extract(first_recv, "^[0-9]{4}")),
         last_recv = as.numeric(str_extract(last_recv, "^[0-9]{4}"))
         ) |>
  mutate(decade_first_recv = floor(first_recv / 10) * 10,
         decade_last_recv = floor(last_recv / 10) * 10) |>
  mutate(decade_last_recv = floor(last_recv / 10) * 10) |>
  replace_na(list(setting = "0 = UNSPECIFIED SETTING",
                  marker = "DO = NOT SPECIFIED OR SEE DESCRIPTION")) 

data_for_plot |>
  pivot_longer(cols = c(decade_first_recv, decade_last_recv),
               names_to = "first_or_last",
               values_to = "decade") |>
  count(first_or_last, decade, marker) |>
  arrange(decade) |>
  mutate(decade = as_factor(decade),
         marker = fct_reorder(marker, n, sum)
  ) |>
  mutate(marker = fct_lump_n(marker, w = n, n = 10)) |>
  filter(marker != "Other") |>
  mutate(marker = fct_rev(marker)) |>
  ggplot(aes(marker, n, fill = decade, group = first_or_last)) +
  geom_col(show.legend = TRUE,
           alpha = 0.7,
           position = position_dodge2(preserve = "single")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_viridis_d(end = 0.9) +
  guides(fill = guide_legend(ncol = 2)) +
  facet_wrap( ~first_or_last) +
  theme(legend.position = c(0.9, 0.6),
        axis.text.x = element_text(angle =75,
                                   hjust = 1)) +
  labs(
    title = "Ten most common marks: first and last recovery",
    subtitle = glue("Count of mark types by decade first and last recovered",
                    "\n{comma(n_marks_counties)} marks in {n_counties} NC counties"),
    x = NULL,
    y = NULL,
    caption = my_caption
  )

```

<br>

## Mark recovery activity

In recent years the last recovered date includes month and day, which reveals the varying amounts of activity in different years and the burstiness (near-vertical lines) in some years (@fig-mark-recovery-activity). Compare years of greatest activity with the densest sections of @fig-mark-condition-age panel A.

```{r, fig.height=8, fig.width=10}
#| label: fig-mark-recovery-activity
#| fig-cap: "Pace of recovery activity"
#| fig-height: 6
#| fig-width: 8

data_for_plot <- nc_marks_counties |>
  st_drop_geometry() |>
  mutate(last_rec_yday = yday(last_recv),
         year = year(last_recv),
         tmp = 1) |>
  filter(year > 2000) |>
  arrange(year, last_rec_yday) |>
    mutate(idx = row_number(),
           cum_mark = cumsum(tmp),
           .by = c(year))

data_for_plot_label <- data_for_plot |>
  filter(cum_mark == max(cum_mark),
         .by = year)

data_for_plot |>
  ggplot(aes(last_rec_yday, cum_mark, color = year, group = year)) +
  geom_line(linewidth = 0.15,
            show.legend = FALSE) + 
  geom_point(size = 0.2,
            show.legend = FALSE) +
  geom_label(data = data_for_plot_label,
            aes(label = year),
            size = 3,
            show.legend = FALSE) +
  labs(
    title = "Pace of recovery activity",
    subtitle = glue("Cumulative by year 2001-2023. {comma(n_marks_counties)} marks in {n_counties} NC counties"),
    x = "Day of the year",
    y = "Cumulative marks recovered in year",
    caption = my_caption
  )

```

<br>

Marks have been recovered by public and private agencies, voluntary groups, and individuals. In each decade since 1980, in these in these `r spell_out(n_counties)` counties, voluntary groups (notably the U.S. Power Squadron) or other geocaching hobbyists have recovered the most marks, the second-most, or both (@fig-last-recovered-by).

```{r, fig.height=10, fig.width=10}
#| label: fig-last-recovered-by
#| fig-cap: "Oranizations ranked by number of recovered survey marks each decade"
#| fig-height: 10
#| fig-width: 10
#| warning: false
#| column: screen-right

nc_marks_counties |>
  st_drop_geometry() |>
  filter(!is.na(last_recv),
         !is.na(last_recby),
         last_recby != "NONE") |>
  mutate(decade = floor(year(last_recv) / 10) * 10) |>
  count(decade, last_recby) |>
  group_by(decade) %>%
  mutate(decade_rank = nrow(cur_data()) - rank(n, ties.method = "first") + 1,
         last_recby_label = glue("{last_recby}: {n}")) |>
  ungroup() |>
  select(decade, last_recby, decade_rank, n, last_recby_label) |>
  ggplot(aes(decade, decade_rank, color = last_recby)) +
  geom_bump(show.legend = FALSE) +
  geom_point(show.legend = FALSE) +
  geom_label(aes(label = last_recby, size = n),
             show.legend = FALSE) +
  scale_x_continuous(position = "top",
                     breaks = 1910 + (0:11) * 10) +
  scale_size_continuous(range = c(2, 6)) +
  scale_y_reverse(breaks = c(1, 5, 10, 15, 20)) +
  scale_color_viridis_d(end = 0.9) +
  labs(
    title = "Oranizations that recovered survey marks",
    subtitle = glue("Ranked by number of marks organization recovered each decade in the selected counties; rank 1 is most common",
                    "\nSize indicates relative number of marks recovered in the decade"),
    x = NULL,
    y = NULL,
    caption = my_caption
  )
    
```

<br>

The U.S. Coast & Geodetic Survey (CGS) is now called the National Geodetic Survey (NGS).^[<https://www.usgs.gov/faqs/how-can-i-find-survey-benchmark-and-triangulation-station-information> ] and is part of the National Ocean Service (still part of NOAA).

@tbl-contrib includes the name associated with each abbreviation used in @fig-last-recovered-by as well as total marks recovered in all decades.

```{r}
#| label: tbl-contrib
#| tbl-cap: "Oranizations that recovered survey marks"

contrib |>
  select(last_recby_org, last_recby, n) |>
  arrange(last_recby_org) |>
  gt() |>
  tab_options(table.font.size = 10) |>
  tab_header(md(glue("**{n_org} organizations recovered survey marks**"))) |>
  tab_source_note(md("*Table: Daniel Moul; source: www.ngs.noaa.gov/cgi-bin/get_contrib2.prl retrieved 2024-02-03*"))

```

<br>

***TODO: add teaser for next chapter***

<br>
